Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-09 _tex/paper2.tex
Built site for gh-pages

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement, and without providing clear guidance on
how to use its measurement scores for hypothesis testing. Specifically,
he cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, as the case imposes the most
extensive set of simplifying assumptions
\citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
this continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not address
how to use the resulting measurement scores for hypothesis testing.
Thus, given these limitations, it is surprising that CJ research has
predominantly relied on Case V to measure different traits, raising
significant concerns about the reliability and validity of such
measurements in contexts where the case's assumptions may not hold
\citep{Kelly_et_al_2022}. Furthermore, although the CJ practice has
attempted to address the issue of hypothesis testing by using the
scores' point estimates or their transformations, a critical question
remains: Is this approach suitable for addressing these inquiries?

@@

SCM~\ref{fig-cj03_scm} presents the first theoretical principles
embedded in the conceptual-population model, which evaluates the impact
of different teaching methods on students' writing ability. This SCM
outlines the relationship between the conceptual-population outcome
\((O^{cp}_{haibjk})\) and several related variables. The subscripts
\(h\) and \(i\) identify the students who authored the texts (i.e., the
individuals). The indices \(a\) and \(b\) represent the texts under
comparison (i.e., the stimuli). The index \(j\) indicates the judge
conducting the comparison, while the index \(k\) captures \emph{repeated
measures designs} \citep[pp.~366-376]{Lawson_2015}, accounting for
experimental conditions where a judge compares the same pair of stimuli
multiple times. Thus, the indexing system supports comparisons between
different texts written by the same student \((h = i;\) \(a \neq b)\)
and between texts written by distinct students \((h \neq i;\) where
\(a = b\) is permitted\()\), each judged once or repeatedly by all
judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\) where \(n_{J}>1\)
and \(n_{K}\geq1)\). However, it excludes cases where a judge compares a
student's text to itself, whether once or multiple times \((h = i;\)
\(a = b;\) \(j = 1,\dots,n_{J};\) \(k = 1,\dots,n_{K};\) where
\(n_{J}>1\) and \(n_{K}\geq1)\), as such comparison lacks practical
relevance within the CJ framework. Here, \(n_{J}\) indicates the total
number of judges, and \(n_{K}\) denotes the number of repeated judgments
each judge performs.

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ha}, T_{ib})\) and their discriminal
difference \((D_{haibjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, based on the arguments developed in
Section~\ref{sec-theory-issue1b} and the recommendations of
\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, the SCM incorporates
the judges' biases \((B_{kj})\). Together with the outcome, these
variables constitute the preliminary set of endogenous variables,
\(V = \{ O_{haibjk}, D_{haibjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
the SCM presents the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\[
\begin{aligned}
  O^{cp}_{haibjk} & := f_{O}(D_{haibjk}) \\
  D_{haibjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
\end{aligned}
\]

@@

Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
SCM~\ref{fig-cj03_scm}, illustrating the expected causal relationships
outlined in Thurstone's theory. The graph shows that the discriminal
processes of the texts \((T_{ha}, T_{ib})\) influence their discriminal
difference \((D_{hiabjk})\), which in turn determines the outcome
\((O^{cp}_{hiabjk})\). It also highlights the influence of judges'
biases \((B_{kj})\) on the discriminal difference. Additionally, the DAG
differentiates between observed endogenous variables, such as the
outcome (solid black circle), and latent endogenous variables, including
the texts' discriminal processes, their discriminal difference, and the
judges' biases (open circles).

@@

Although specifying a data structure is not mandatory when using SCMs
and DAGs, defining one in this case can improve clarity and facilitate
the description of the system. Thus, to re-express the scalar form of
the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
vectorized form, we first define the vectors \(I\) and \(J\), along with
the matrices \(IA\) and \(JK\), as follows:
\begin{equation}\phantomsection\label{eq-mat11}{
I = \begin{bmatrix}
1 \\
\vdots \\
h \\
\vdots \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A}
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
\vdots & \vdots \\
n_{J} & n_{K}
\end{bmatrix}
}\end{equation}

Here, each element of \(I\) represents a unique individual \(h\) or
\(i\), where \(n_{I}\) denotes the total number of individuals.
Similarly, each element of \(J\) corresponds to a unique judge \(j\),
with \(n_{J}\) indicating the total number of judges. Moreover, each row
of \(IA\) represents a unique pairing of individuals \(h, i\) with
stimuli \(a, b\). As a result, the matrix \(IA\) contains
\(n_{I} \cdot n_{A}\) rows and \(2\) columns, where \(n_{A}\) specifies
the number of stimuli available per individual. Likewise, each row of
\(JK\) associates a judge \(j\) with a (repeated) judgment index \(k\).
Consequently, the matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\)
columns, where \(n_{K}\) indicates the number of repeated judgments each
judge makes.

@@

\begin{equation}\phantomsection\label{eq-mat12}{
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & a & i & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}
\end{bmatrix}
}\end{equation}

@@

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{\emph{Relevant variables} are
  those that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
variables \((X_{I})\), such as teaching method, and students'
idiosyncratic errors \(e_{I}\) causally influence the latent variable
representing students' writing-quality trait \((T_{I})\). The error term
\(e_{I}\) captures variations in students' traits unexplained by
\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
\(q_{I}\) columns (variables), and both \(e_{I}\) and \(T_{I}\) are
latent vectors of length \(n_{I}\). Additionally, this branch shows how
\(T_{I}\), along with \emph{relevant} \footnote{refer to footnote 2.}
text-related variables \((X_{IA})\) (e.g., text length), and texts'
idiosyncratic errors \(e_{IA}\) causally influence the texts'
writing-quality trait \((T_{IA})\), the first primary latent variable of
interest. The error term \(e_{IA}\) captures variations in the texts'
traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
rows and \(q_{IA}\) columns (variables), while \(e_{IA}\) and \(T_{IA}\)
are latent matrices with \(n_{I}\) rows and \(n_{A}\) columns.

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 2.}
judge-related variables \((Z_{J})\), such as judgment expertise, and
judges' idiosyncratic errors \(e_{J}\) causally influence the latent
variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
captures variations in judges' bias unexplained by \(Z_{J}\). Here,
\(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
columns (variables), and both \(e_{J}\) and \(B_{J}\) are latent vectors
of length \(n_{J}\). Furthermore, the branch shows how \(B_{J}\), along
with \emph{relevant} \footnote{refer to footnote 2.} judgment-related
variables \((Z_{JK})\) (e.g., the number of judgments a judge makes),
and judgments' idiosyncratic errors \(e_{JK}\) causally influence the
biases associated with each text \(B_{JK}\), the second primary latent
variable of interest. The error \(e_{JK}\) captures variations in
judgments unexplained by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{JK}\) is an
observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\)
columns (variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent
matrices with \(n_{J}\) rows and \(n_{K}\) columns

@@

To incorporate the sampling mechanism and facilitate the interpretation
of the sample-comparison model, we first define the \emph{data sampling
process} using the binary vector variables \(S_{I}\), \(S_{J}\),
\(S_{IA}\), and \(S_{JK}\) as follows:
\begin{equation}\phantomsection\label{eq-mat21}{
S_{I} = \begin{bmatrix}
i_{(1)} \\
\vdots \\
i_{(h)} \\
\vdots \\
i_{(i)} \\
\vdots \\
i_{(nI)}
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
j_{(1)} \\
\vdots \\
j_{(j)} \\
\vdots \\
j_{(nJ)}
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
ia_{(1,1)} \\
\vdots \\
ia_{(1,n_{A})} \\
\vdots \\
ia_{(h,a)} \\
\vdots \\
ia_{(i,b)} \\
\vdots \\
ia_{(nI,1)} \\
\vdots \\
ia_{(nI,nA)}
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
jk_{(1,1)} \\
\vdots \\
jk_{(1,n_{K})} \\
\vdots \\
jk_{(j,k)} \\
\vdots \\
jk_{(nJ,1)} \\
\vdots \\
jk_{(nJ,nK)}
\end{bmatrix}
}\end{equation}

@@

Finally, we define the \emph{sample mechanism} \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). Each element \(s_{(h,a,i,b,j,k)}\) is a binary value
indicating the presence or absence of data rows in the matrix \(R\)
resulting from the sample mechanism, as in equation (\ref{eq-mat26}).
Thus, the vector contains \(n\) elements, matching the number of rows in
\(R\), and the sum of its elements represents the total data sample:
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals.

\begin{equation}\phantomsection\label{eq-mat25}{
S = \begin{bmatrix}
s_{(1,1,1,2,1,1)} \\
\vdots \\
s_{(1,1,1,2,1,n_{K})} \\
\vdots \\
s_{(h,a,i,b,j,k)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

\begin{equation}\phantomsection\label{eq-mat26}{
s_{(h,i,a,b,j,k)} = \begin{cases} 
1 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are sampled} \\
0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
\end{cases}
}\end{equation}

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
directly access this outcome due to the sampling mechanism. The DAG also
depicts the \emph{sample design} vector \(S\) as a causal factor
influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
encloses this vector, indicating that it is a conditioned variable.
Here, \emph{conditioning} means that researchers restrict their focus to
the elements of \(O^{cp}_{R}\) that satisfy \(s_{(h,i,a,b,j,k)}=1\)
\citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
vector selecting \emph{all repeated judgments a subset of judges makes
for the subset of stimuli produced by the sampled individuals}. This
process generates the sample-comparison outcome \(O^{sc}_{R}\).

@@

As in the previous section, to facilitate the interpretation of the
sample-comparison model, we first define the \emph{comparison mechanism}
using the binary vector variable \(C\) as follows:
\begin{equation}\phantomsection\label{eq-mat27}{
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(h,a,i,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

Where \(C\) contains \(n\) elements corresponding to the number of rows
in the \(R\) matrix, with each element \(c_{(h,a,i,b,j,k)}\) being a
binary value indicating the presence or absence of data rows in the
\(R\) similar to equation (\ref{eq-mat26}). This definition assumes that
the sampled judges \emph{do not} perform all repeated judgments for the
subset of stimuli produced by the sampled individuals. Instead, they
conduct a ``sufficient number of comparisons'' \(n_{C}\) to ensure the
accurate estimation of the proportion \(P(B>A)\) for each pair of
stimuli \citep[pp.~267]{Thurstone_1927b}.

DAG~\ref{fig-cj14_dag} also incorporates the comparison mechanism \(C\)
into the conceptual-population model. The DAG represents the
sample-comparison outcome \(O^{sc}_{R}\) as unobserved, highlighting
that researchers cannot directly access this outcome due to the
comparison mechanism. Moreover, it represents the \emph{comparison
mechanism} \(C\) as a causal factor influencing the final comparative
outcome \(O_{R}\). Equivalent to the sampling mechanism, a square
encloses this vector, indicating that \(C\) is a conditioned variable
that determines \emph{which repeated judgments the subset of judges
makes for the subset of stimuli produced by the sampled individuals}.
This process ultimately derives the final comparative outcome \(O_{R}\).

Notably, the DAG shows that \(C\) is also independent of all other
variables in the model. This independence implies that the conceptual
model represented by DAG~\ref{fig-cj14_dag} applies exclusively to
Random Allocation Comparative Designs \citep{Bramley_2015} or Incomplete
Block Designs \citep{Lawson_2015}. In these designs, every repeated
judgment has an equal probability of being included in the sample.

>>>>

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement, and without providing clear guidance on
how to use its measurement scores for hypothesis testing. Specifically,
he cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, as the case imposes the most
extensive set of simplifying assumptions
\citep{Bramley_2008, Kelly_et_al_2022} (see
Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's primary
goal was to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
this continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not address
how to use the resulting measurement scores for hypothesis testing.
Thus, given these limitations, it is surprising that CJ research has
predominantly relied on Case V to measure different traits, raising
significant concerns about the reliability and validity of such
measurements in contexts where the case's assumptions may not hold
\citep{Kelly_et_al_2022, Andrich_1978}. Furthermore, although the CJ
practice has attempted to address the issue of hypothesis testing by
using the scores' point estimates or their transformations, a critical
question remains: Is this approach suitable for addressing these
inquiries?

@@

SCM~\ref{fig-cj03_scm} presents the first theoretical principles
embedded in the conceptual-population model, which evaluates the impact
of different teaching methods on students' writing ability. This SCM
outlines the relationship between the conceptual-population outcome
\((O^{cp}_{iahbjk})\) and several related variables. The subscripts
\(i\) and \(h\) identify the students who authored the texts (i.e., the
individuals). The indices \(a\) and \(b\) represent the texts under
comparison (i.e., the stimuli). The index \(j\) indicates the judge
conducting the comparison, while the index \(k\) captures \emph{repeated
measures designs} \citep[pp.~366-376]{Lawson_2015}, accounting for
experimental conditions where a judge compares the same pair of stimuli
multiple times. Thus, the indexing system supports comparisons between
different texts written by the same student \((i = h;\) \(a \neq b)\)
and between texts written by distinct students \((i \neq h;\) where
\(a = b\) is permitted\()\), each judged once or repeatedly by all
judges \((j = 1,\dots,n_{J};\) \(k = 1,\dots,n_K;\) where \(n_{J}>1\)
and \(n_{K}\geq1)\). However, it excludes cases where a judge compares a
student's text to itself, whether once or multiple times \((i = h;\)
\(a = b;\) \(j = 1,\dots,n_{J};\) \(k = 1,\dots,n_{K};\) where
\(n_{J}>1\) and \(n_{K}\geq1)\), as such comparison lacks practical
relevance within the CJ framework. Here, \(n_{J}\) indicates the total
number of judges, and \(n_{K}\) denotes the number of repeated judgments
each judge performs.

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, based on the arguments developed in
Section~\ref{sec-theory-issue1b}, the SCM incorporates the judges'
biases \((B_{kj})\). Together with the outcome, these variables
constitute the preliminary set of endogenous variables,
\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\). Finally,
the SCM presents the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\[
\begin{aligned}
  O^{cp}_{iahbjk} & := f_{O}(D_{iahbjk}) \\
  D_{iahbjk} & := f_{D}(T_{ia}, T_{hb}, B_{jk})
\end{aligned}
\]

@@

Figure~\ref{fig-cj03_dag} displays the DAG corresponding to
SCM~\ref{fig-cj03_scm}, illustrating the expected causal relationships
outlined in Thurstone's theory. The graph shows that the discriminal
processes of the texts \((T_{ia}, T_{hb})\) influence their discriminal
difference \((D_{iahbjk})\), which in turn determines the outcome
\((O^{cp}_{iahbjk})\). It also highlights the influence of judges'
biases \((B_{kj})\) on the discriminal difference. Additionally, the DAG
differentiates between observed endogenous variables, such as the
outcome (solid black circle), and latent endogenous variables, including
the texts' discriminal processes, their discriminal difference, and the
judges' biases (open circles).

@@

Although specifying a data structure is not mandatory when using SCMs
and DAGs, defining one in this case can improve clarity and facilitate
the description of the system. Thus, to re-express the scalar form of
the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
vectorized form, we first define the vectors \(I\) and \(J\), along with
the matrices \(IA\) and \(JK\), as follows:
\begin{equation}\phantomsection\label{eq-mat11}{
I = \begin{bmatrix}
1 \\
\vdots \\
i \\
\vdots \\
h \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
i & a \\
\vdots & \vdots \\
h & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A}
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
\vdots & \vdots \\
n_{J} & n_{K}
\end{bmatrix}
}\end{equation}

Here, each element of \(I\) represents a unique individual \(i\) or
\(h\), where \(n_{I}\) denotes the total number of individuals.
Similarly, each element of \(J\) corresponds to a unique judge \(j\),
with \(n_{J}\) indicating the total number of judges. Moreover, each row
of \(IA\) represents a unique pairing of individuals \(i, h\) with
stimuli \(a, b\). As a result, the matrix \(IA\) contains
\(n_{I} \cdot n_{A}\) rows and \(2\) columns, where \(n_{A}\) specifies
the number of stimuli available per individual. Likewise, each row of
\(JK\) associates a judge \(j\) with a (repeated) judgment index \(k\).
Consequently, the matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\)
columns, where \(n_{K}\) indicates the number of repeated judgments each
judge makes.

@@

\begin{equation}\phantomsection\label{eq-mat12}{
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
i & a & h & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}
\end{bmatrix}
}\end{equation}

@@

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{\emph{Relevant variables} are
  those that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
variables \((X_{I})\), such as teaching method, and students'
idiosyncratic errors \(e_{I}\) causally influence the latent variable
representing students' writing-quality trait \((T_{I})\). The error term
\(e_{I}\) captures variations in students' traits unexplained by
\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
\(q_{I}\) columns (variables), and both \(e_{I}\) and \(T_{I}\) are
latent vectors of length \(n_{I}\). Additionally, this branch shows how
\(T_{I}\), along with \emph{relevant} \footnote{refer to footnote 2.}
text-related variables \((X_{IA})\) (e.g., text length), and texts'
idiosyncratic errors \(e_{IA}\) causally influence the texts'
written-quality trait \((T_{IA})\), the first primary latent variable of
interest. The error term \(e_{IA}\) captures variations in the texts'
traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
rows and \(q_{IA}\) columns (variables), while \(e_{IA}\) and \(T_{IA}\)
are latent matrices with \(n_{I}\) rows and \(n_{A}\) columns.

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 2.}
judge-related variables \((Z_{J})\), such as judgment expertise, and
judges' idiosyncratic errors \(e_{J}\) causally influence the latent
variable representing judges' bias \((B_{J})\). The error \(e_{J}\)
captures variations in judges' bias unexplained by \(Z_{J}\). Here,
\(Z_{J}\) is an observed matrix with \(n_{J}\) rows and \(q_{J}\)
columns (variables), and both \(e_{J}\) and \(B_{J}\) are latent vectors
of length \(n_{J}\). Furthermore, the branch shows how \(B_{J}\), along
with \emph{relevant} \footnote{refer to footnote 2.} judgment-related
variables \((Z_{JK})\) (e.g., the number of judgments a judge makes),
and judgments' idiosyncratic errors \(e_{JK}\) causally influence the
judges' biases associated with each text \(B_{JK}\), the second primary
latent variable of interest. The error \(e_{JK}\) captures variations in
judgments unexplained by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{JK}\) is an
observed matrix with dimension \(n_{J} \cdot n_{K}\) rows and \(q_{JK}\)
columns (variables), while \(e_{JK}\) and \(B_{JK}\) are latent latent
matrices with \(n_{J}\) rows and \(n_{K}\) columns

@@

To incorporate the sampling mechanism and facilitate the interpretation
of the sample-comparison model, we first define the \emph{data sampling
process} using the binary vector variables \(S_{I}\), \(S_{J}\),
\(S_{IA}\), and \(S_{JK}\) as follows:
\begin{equation}\phantomsection\label{eq-mat21}{
S_{I} = \begin{bmatrix}
i_{(1)} \\
\vdots \\
i_{(i)} \\
\vdots \\
i_{(h)} \\
\vdots \\
i_{(nI)}
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
j_{(1)} \\
\vdots \\
j_{(j)} \\
\vdots \\
j_{(nJ)}
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
ia_{(1,1)} \\
\vdots \\
ia_{(1,n_{A})} \\
\vdots \\
ia_{(i,a)} \\
\vdots \\
ia_{(h,b)} \\
\vdots \\
ia_{(nI,1)} \\
\vdots \\
ia_{(nI,nA)}
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
jk_{(1,1)} \\
\vdots \\
jk_{(1,n_{K})} \\
\vdots \\
jk_{(j,k)} \\
\vdots \\
jk_{(nJ,1)} \\
\vdots \\
jk_{(nJ,nK)}
\end{bmatrix}
}\end{equation}

@@

Finally, we define the \emph{sample mechanism} \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). Each element \(s_{(i,a,h,b,j,k)}\) is a binary value
indicating the presence or absence of data rows in the matrix \(R\)
resulting from the sample mechanism, as in equation (\ref{eq-mat26}).
Thus, the vector contains \(n\) elements, matching the number of rows in
\(R\), and the sum of its elements represents the total data sample:
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals.

\begin{equation}\phantomsection\label{eq-mat25}{
S = \begin{bmatrix}
s_{(1,1,1,2,1,1)} \\
\vdots \\
s_{(1,1,1,2,1,n_{K})} \\
\vdots \\
s_{(i,a,h,b,j,k)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

\begin{equation}\phantomsection\label{eq-mat26}{
s_{(i,a,h,b,j,k)} = \begin{cases} 
1 & \text{if data elements } i,a,h,b,j,k \text{ from } R \text{ are sampled} \\
0 & \text{if data elements } h,i,a,b,j,k \text{ from } R \text{ are missing}
\end{cases}
}\end{equation}

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet{McElreath_2020} and \citet{Deffner_et_al_2022},
DAG~\ref{fig-cj14_dag} represents the conceptual-population outcome
\(O^{cp}_{R}\) as unobserved, emphasizing that researchers cannot
directly access this outcome due to the sampling mechanism. The DAG also
depicts the \emph{sample design} vector \(S\) as a causal factor
influencing the sample-comparison outcome \(O^{sc}_{R}\). A square
encloses this vector, indicating that it is a conditioned variable.
Here, \emph{conditioning} means that researchers restrict their focus to
the elements of \(O^{cp}_{R}\) that satisfy \(s_{(i,a,h,b,j,k)}=1\)
\citep{Neal_2020, McElreath_2020}. Essentially, \(S\) serves as the
vector selecting \emph{all repeated judgments a subset of judges makes
for the subset of stimuli produced by the sampled individuals}. This
process generates the sample-comparison outcome \(O^{sc}_{R}\).

@@

As in the previous section, we begin defining the \emph{comparison
mechanism} using the binary vector variable \(C\) to facilitate the
interpretation of the sample-comparison model (see equation
(\ref{eq-mat27})). The vector contains \(n\) elements corresponding to
the number of rows in the \(R\) matrix, with each element
\(c_{(i,a,h,b,j,k)}\) being a binary value indicating the presence or
absence of data rows in \(R\), a definition similar to that of \(S\) in
equation (\ref{eq-mat26}).
\begin{equation}\phantomsection\label{eq-mat27}{
C = \begin{bmatrix}
c_{(1,1,1,2,1,1)} \\
\vdots \\
c_{(1,1,1,2,1,n_{K})} \\
\vdots \\
c_{(i,a,h,b,j,k)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
c_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

The DAG~\ref{fig-cj14_dag} also integrates the comparison mechanism
\(C\) into the conceptual-population model, which ultimately derives the
final comparative outcome \(O_{R}\). First, it shows the
sample-comparison outcome \(O^{sc}_{R}\) as unobserved, highlighting
that researchers cannot directly access this outcome due to the
comparison mechanism. Second, it depicts the \emph{comparison mechanism}
\(C\) as a causal factor influencing the final comparative outcome
\(O_{R}\). As with the sampling mechanism, a square encloses \(C\) to
indicate that it is a conditioned variable--that is, one that determines
\emph{which repeated judgments the subset of judges makes for the subset
of stimuli produced by the sampled individuals}. In essence, \(C\)
reflects the assumption that judges \emph{do not} perform all possible
repeated judgments for the selected stimuli. Instead, they carry out a
``sufficient number of comparisons'' \(n_{C}\) to allow for accurate
estimation of the proportion \(P(B>A)\) for each stimulus pair
\citep[pp.~267]{Thurstone_1927b}.

Notably, DAG~\ref{fig-cj14_dag} shows that \(C\) is independent of all
other variables in the model. This independence implies that the
conceptual model represented by the DAG applies exclusively to Random
Allocation Comparative Designs \citep{Bramley_2015} or Incomplete Block
Designs \citep{Lawson_2015}. In these designs, every repeated judgment
has an equal probability of being included in the sample.
